---
title: 'Polling Data Analysis'
author: "Hemanth Bharatha Chakravarthy"
date: "2/23/2019"
output: html_document
---

```{r setup, include=FALSE}
# Problem Set 4, Hemanth Bharatha Chakravarthy
# Because there are many plots here, the default option of 
# echo = FALSE is being set
knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE)
# Including the relevant libraries
library(tidyverse)
library(ggplot2)
library(ggthemes)
# I'm using default Tidyverse styles here
library(styler)
library(lintr)
library(ggplot2)
library(readr)
library(dplyr)
library(knitr)
# This package helps read our .csv file
library(readr)
library(janitor)
# These are a part of the visualization/stylistic choices for this project
library(ggridges)
library(viridis)
# Some useful computation is in lubridate
library(lubridate)
library(devtools)
library(gt)
# This project uses Harvard enrollment data obtained from the Registrar of the FAS
# The .xlsx file can be found on Canvas
# Loading the excel file skipping so that the titles of each column match those in the R df
# Cleaning names
polls <- read_csv("ps_4_elections-poll-nc09-3.csv", 
                  col_types = cols(
  .default = col_character(),
  turnout_scale = col_double(),
  turnout_score = col_double(),
  w_LV = col_double(),
  w_RV = col_double(),
  final_weight = col_double(),
  timestamp = col_datetime(format = "")
  )) %>% 
  clean_names()
# Note on style: I'm using traditional Tidyverse style and 
# This is auto-checked by `lintr::lint("test.R")````
```


# 1 Distribution of Votes Based on Race
```{r table_q1}
# Question 2
# Create new tibble
race <- polls %>% 
  select(race_eth, response, final_weight) %>% 
  # Filter out nonsense rows
  filter(!is.na(response), 
        race_eth != "[DO NOT READ] Don't know/Refused",
        response %in% c("Dem", "Rep", "Und", "3")
        ) %>%  
  group_by(race_eth, response) %>% 
  # SUmmarize by weight because we're adjusting for likelihood to turnout
  # Take sum of weights
  summarize(total = sum(final_weight)) %>%
  # Rearrange table to make dem, rep, etc columns
  spread(key = response, value = total) %>%
  # set NA as 0
  mutate_all(funs(replace(., is.na(.), 0))) %>% 
  ungroup() %>% 
  group_by(race_eth) %>% 
  # Mutate a total row
  mutate(y = Dem + Rep + Und + `3`) %>% 
  # COnvert to %s
  mutate(Dem = (Dem)/y,
         Rep = (Rep)/y,
         Und = (Und)/y) %>% 
  ungroup() %>% 
  # Select required rows
  select(race_eth, Dem, Rep, Und) %>% 
  # Reorder in given order of races
  slice(match(c("White", "Black", "Hispanic", "Asian", "Other"), race_eth))
  
# Table 
# Use gt_tbl
race %>%
  gt(rowname_col = "race_eth") %>%
  # header and source
  tab_header(title = "Polling Results in North Carolina 9th Congressional District") %>%
  tab_source_note(
    source_note = "Data from The New York Times Upshot"
  ) %>% 
  # Relabelling column
  # TO readers not familiar with the US, the full party names would be useful
  # Furthermore, UND. isn't directly understood as undecided by lay readers
  # So, I made the design decision to include the full names
  cols_label(
    Dem = html("Democrat"),
    Rep = html("Republican"),
    Und = html("Undecided")
  ) %>%
  # Formatting as percent with 0 decimal points shown and % sign
  fmt_percent(
  columns = vars(Dem),
  decimals = 0) %>%
  fmt_percent(
  columns = vars(Rep),
  decimals = 0) %>%
  fmt_percent(
  columns = vars(Und),
  decimals = 0)
```

# 2 Education vs Turnout Weights 
Turnout weights measured likelihood to turnout to vote.

```{r turnout_education_plot}
# Q3
# Create new tibble with education data
educ_graph <- polls %>% 
  # Filter out nonsense rows
  filter(!is.na(response), 
        educ != "[DO NOT READ] Refused") %>% 
  select(educ, final_weight)  
# Plot
educ_graph %>% 
  # X axis is discrete with education for now
  # We will flip this later 
  # But, this is conventional and intuitive
  ggplot(aes(x = educ, y = final_weight)) +
  # First plot is a violin plot with width adjusted to approx match the given plot
  geom_violin(width = 1) + 
  coord_flip() +
  # Jitter plot with design changed to approx match the given plot
  geom_jitter(alpha=0.6, size=0.8, width = 0.26) +
  # Labels and no X label
  labs(
    title = "More Educated Matter Less in North Carolina 9th",
    subtitle = "Poll gives more weight to people who are more likely to participate in polls", 
    caption = "New York Times Upshot/Siena College 2018 live polls") +
  ylab("Weight Given to Respondent in Calculating Poll Results") +
  xlab(NULL) 
```

# 3 Race, Gender vs Voting Republican
The purpose of this plot is to study whether gender was a key factor of voting in the 9th District. However, it appears that while race playaed a great role, except for Hispanic women who were somewhat less likely to vote Republican than Hispanic men, other races showed insignificant gender-based difference.

**This is interesting because it studies whether within the same racial background, gender could swing votes, but we find that it did not do so.**


```{r gender_race_plot}
# The purpose of this plot is to study whether gender was a key 
# factor of voting in the 9th District. 
# This is interesting because it studies whether within the same racial background, gender could swing votes
#  However, it appears that while race playaed a great role, except
# for Hispanic women who were somewhat less likely to vote Republican than
# Hispanic men, other races showed insignificant gender-based difference.
# Create new tibble
gender_race <- polls %>% 
  # Filter out nonsense values
  filter(!is.na(response), 
        race_eth != "[DO NOT READ] Don't know/Refused",
        response %in% c("Dem", "Rep", "Und", "3"),
        !is.na(gender)
        ) %>%  
  # Find the total turnout adjusted votes based on gender and race
  group_by(gender, race_eth, response) %>% 
  summarize(total = sum(final_weight)) %>%
  spread(key = response, value = total) %>%
  # set NA as 0
  mutate_all(funs(replace(., is.na(.), 0))) %>% 
  ungroup() %>% 
  group_by(gender) %>% 
  # Select required rows
  select(gender, race_eth, Rep) 
# Plot
gender_race %>% 
  ggplot(aes(x = gender, y = Rep, color = gender)) +
  # DESIGN JUSTIFICATION for bar graph:
  # Given that we're comparing total weights, a bar graph is most suitable
  # Bar graphs are also intuitive and wasy to read
  # With the need to facet wrap, bar graph becomes more ideal
  # We don't need legend because the axis is self explanatory
  geom_col(show.legend = FALSE) +
  # compare by internal gender differences between races
  facet_wrap(~race_eth) +
  # use log scale given uneven distribution of frequencies
  scale_y_log10() +
  # Add labels
  labs(title = "White Women Also Voted Republican in North Carolina 9th Dist.",
       subtitle = "Of Republican respondents, adjusting for turnout, there was little difference based on gender",
       caption = "Data from NY Times Upshot") +
  xlab("Gender of Respondent") +
  ylab("Turnout Adjusted Republican Votes") +
  # Using the stata theme to make it look more attractive and interesting
  # It also makes neat boxlike structures and cleans up the facet wrap
  theme_stata()
```
